---
title: "Dose worlds: when the amount matters more than the assignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose worlds: when the amount matters more than the assignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why a continuous treatment axis

Not every psychotherapy question is a two-arm contrast. How many
sessions a patient attends, how much alliance exposure they accumulate,
how strongly a pairing matches: these are questions about *amount*, and
forcing them into a binary treatment throws the question away before
the analysis starts.

Since version 0.2.0, `sim_stream(z_type = "dose")` draws a continuous
exposure `z` uniform on [0, 1] instead of a binary arm. Everything else
about the world stays the same: nesting, sessions, dropout, measurement
error. All effect parameters act *per unit dose*, so `tau = 0.5` means:
moving a patient from dose 0 to dose 1 shifts the outcome by 0.5.

## Build a dose world

```{r}
library(windkanal)

s <- sim_stream(n_therapists = 20, patients_per_therapist = 10,
                n_sessions = 4, z_level = "patient",
                z_type = "dose",
                tau = 0.5, tau_x = 0.5, icc = 0.10, seed = 1)
range(s$z)
head(s[, c("therapist_id", "patient_id", "session", "z", "x", "score")])
```

`z_level` keeps its meaning: `"therapist"` gives every patient of one
therapist the same dose (a caseload-level exposure), `"patient"` varies
the dose within therapists.

Binary worlds are untouched: `z_type = "binary"` remains the default
and reproduces earlier versions bit for bit under the same seed.

## The same statistical trap, on a slope

The point of a dose world is that the nesting problem does not go away
when the treatment becomes continuous; it moves into the slope. A naive
regression of outcome on dose treats every session row as independent
evidence:

```{r}
p <- patients(s)                      # one row per patient
naiv <- lm(score_mean ~ z, data = p)
coef(summary(naiv))["z", , drop = FALSE]
```

The point estimate is fine in expectation; the standard error is the
casualty, exactly as in the binary case, because patients still share
therapists and sessions still share patients. A mixed model with a
therapist intercept prices that dependence in. On a single world the
two intervals just look different; run the pair over many seeds and the
naive interval covers the truth far below its nominal rate, which is
the package's core demonstration (see the main vignette).

## Guard rails

The arm-based CATE machinery (`fit_cate_*()`, `fit_het_grf()`) is
built for binary contrasts. Handing it a dose world quietly would
reinterpret a continuous `z` as two arms; windkanal refuses instead:

```{r, error = TRUE}
fit_cate_pai(p)
```

This is deliberate. A dose world answers dose questions (regression
slopes, dose-response shapes); the binary CATE toolbox answers contrast
questions. The error keeps the two from blending silently.

## Where this is going

The dose axis is the first step of a longer arc: treatment courses as
processes rather than assignments, with recovery kinetics and a
session-wise alliance as explicit model blocks. The public roadmap
lives in the repository's issues.
