---
title: "windkanal: test your tools before they meet real patients"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{windkanal: test your tools before they meet real patients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The problem this package solves

Suppose you want to know whether a statistical tool keeps its promises
on psychotherapy data: patients nested in therapists, outcomes arriving
session by session, dropout, measurement error. On real data you can
never check this, because the true effect is unknown. windkanal builds
clinics in which the truth is known by construction, so any claim an
estimator makes can be compared against what was actually put into the
world.

This vignette walks through one full cycle: build a world, let two
estimators disagree, and turn that disagreement into a measured error
rate.

## Build a clinic

`sim_stream()` returns one row per attended session. Here: 20
therapists with 10 patients each, four sessions, treatment assigned at
the patient level, a true average effect of 0.5, and a therapist share
of outcome variance of 10 percent.

```{r}
library(windkanal)

s <- sim_stream(n_therapists = 20, patients_per_therapist = 10,
                n_sessions = 4, z_level = "patient",
                tau = 0.5, tau_x = 0.5, icc = 0.10, seed = 1)
head(s)
```

Every world is fully determined by its seed. Rerunning the call with
`seed = 1` reproduces this clinic bit for bit, which is why `seed` is a
required argument everywhere in the package.

## The founding observation

Analyze the same world twice. The naive regression treats all session
rows as independent; the mixed model knows that sessions sit inside
patients and patients sit inside therapists.

```{r}
fit_z_naive(s)
fit_z_satt(s)
```

Both estimates are close to each other, but the naive standard error is
far smaller: the model counts 800 pieces of information where it truly
has far fewer. In this world the truth is 0.5; check for yourself
which of the two intervals contains it.

## From anecdote to error rate

One world proves nothing. `mc_run()` repeats the question over many
seeded worlds; here, worlds in which the treatment does nothing at all,
assigned at the therapist level. An honest tool should raise a false
alarm in about five percent of them.

```{r}
mc <- mc_run(n_reps = 30,
             sim_args = list(n_therapists = 20, patients_per_therapist = 10,
                             n_sessions = 4, z_level = "therapist", tau = 0),
             fit_fns = list(naive = fit_z_naive, satt = fit_z_satt),
             seed_start = 1)
mc_summary(mc, truth = 0)
```

The pattern that motivates the package appears already at 30 worlds:
the naive analysis raises false alarms at many times its nominal rate,
while the Satterthwaite mixed model stays near five percent. (For
publication-grade numbers, run hundreds of worlds; this chunk is kept
small so the vignette builds quickly.)

## Person-specific effects

The same interface serves estimators of individual treatment effects.
These calls need the suggested packages grf and stochtree and take
longer, so they are shown but not run here:

```{r, eval = FALSE}
g <- fit_cate_grf(s, num_trees = 1000, seed = 1)   # causal forest
b <- fit_cate_bcf_ml(s, nburn = 500, nsim = 500,   # multilevel BCF
                     rfx = TRUE, seed = 1)
head(b$tau_hat)      # one effect estimate per patient
attr(b, "ate")       # average effect with posterior interval
```

Because `sim_stream()` also wrote the true individual effects into the
world, you can correlate `tau_hat` with the truth, count how often the
per-person intervals cover it, and decide which tool you would trust in
your own setting.

## Calibrated presets, including your own

A preset is a claim about reality: a named set of `sim_stream()`
arguments in which every number carries a citation and a status.

```{r}
p <- preset("ambulanz_de")
attr(p, "sources")[, c("parameter", "wert", "status")]
```

You can define presets for your own clinic or country in the same
form: a named list of `sim_stream()` arguments with a `sources` data
frame attached. Values you cannot calibrate are labelled open rather
than silently guessed; sample sizes stay out of presets deliberately,
because how large your simulated clinic is remains a design choice,
not a reality claim.

```{r}
meine_klinik <- list(n_sessions = 12, icc = 0.08, dropout = 0.02)
attr(meine_klinik, "sources") <- data.frame(
  parameter = c("n_sessions", "icc", "dropout"),
  quelle    = c("clinic records", "own MLM estimate", "open"),
  status    = c("belegt", "belegt", "offen"))

s2 <- do.call(sim_stream, c(meine_klinik,
        list(n_therapists = 15, patients_per_therapist = 8,
             tau = 0.3, seed = 2)))
nrow(s2)
```

## Where to go from here

`snapshot()` and `replay()` freeze and replay growing data streams, so
you can compare analysis habits (continuous peeking versus
preregistered gates, `run_peek()` and `run_gates()`) on identical
histories. `sim_items()` and `scale_scores()` take the simulation down
to the questionnaire item level, with reliability as a measured
attribute. And every function help page documents which part of the
data-generating equation it touches.

If you test a tool on your own preset and it breaks: that is not a
failure of the exercise. Better here than in a clinic.
